---
title: "Analysis with adegenet"
author: "Denise Martini"
date: "12/14/2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Using the adegenet package in R  to run some tests on my GBS dataset. Hopefully some interesting and nice plots as well. 
Loading all the packages I need.

```{r packages}
library(adegenet)
library(ape)
library(RColorBrewer)
```

I prepared the input in the pop_structure directory, they are `.raw` Plink files, that you can obtain with the PLINK `--recode A` command.
Importing the input file:

```{r input}
GBS <- read.PLINK(file="../pop_structure/maxmiss90_common_snps.raw", map.file = "../pop_structure/maxmiss90_common_snps.map")
```

Tha reminds me that `parallel` is also a required package. Luckily enough, adegenet look for it and loads it on its own. A few checks on the dataset before stary=ting:

```{r checks, fig.height=5, fig.width=7, out.width=850px}
# checking that population and individual names are fine
GBS@pop
indNames(GBS)
# checking missing data and allele frequency distributions
glPlot(GBS)
```

Everything as expected. 




